---
title: "CanvasXpress in Quarto (R)"
subtitle: "Interactive charts straight from R code chunks"
author: "CanvasXpress"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    page-layout: full
---

## What this is

This is the **R-native** companion to [`demo.qmd`](demo.qmd). Instead of the
`canvasxpress` shortcode + JSON specs, every chart here is produced by an R code chunk
using the [`canvasXpress`](https://cran.r-project.org/package=canvasXpress) package — an
htmlwidget. In RStudio each chunk renders directly into the **Viewer** pane as you run it
(⌘⇧⏎ / Ctrl+Shift+Enter), and again in the knitted HTML.

No shortcode, no `_extensions`, no CDN — the widget bundles the library.

``` r
install.packages("canvasXpress")   # once
```

```{r}
#| label: setup
#| include: false
library(canvasXpress)
```

## Bar chart

`canvasXpress` takes a matrix whose row names are the variables (series) and column names
the samples:

```{r}
#| label: bar
y <- matrix(c(10, 14, 9, 17,
              6,  7,  5,  9),
            nrow = 2, byrow = TRUE,
            dimnames = list(c("Revenue", "Cost"),
                            c("Q1", "Q2", "Q3", "Q4")))

canvasXpress(data = y,
             graphType   = "Bar",
             title       = "Quarterly Revenue vs Cost",
             graphOrientation = "vertical",
             showLegend  = TRUE)
```

## Scatter plot

Points are the rows; the columns are the axis variables. `sizeBy` maps a third column to
bubble size:

```{r}
#| label: scatter
cars <- mtcars[1:10, c("wt", "mpg", "hp")]

canvasXpress(data = cars,
             graphType  = "Scatter2D",
             title      = "Weight vs. MPG (bubble = horsepower)",
             xAxis      = list("wt"),
             yAxis      = list("mpg"),
             xAxisTitle = "Weight (1000 lbs)",
             yAxisTitle = "Miles per gallon",
             sizeBy     = "hp")
```

## Heatmap

Clustering is done by CanvasXpress at render time:

```{r}
#| label: heatmap
set.seed(1)
m <- matrix(rnorm(30), nrow = 5,
            dimnames = list(paste0("Gene", 1:5), paste0("S", 1:6)))

canvasXpress(data = m,
             graphType         = "Heatmap",
             title             = "Expression Heatmap",
             samplesClustered  = TRUE,
             variablesClustered = TRUE)
```

## Boxplot

A single measurement variable split by a categorical factor. The factor is passed as a
sample annotation (`smpAnnot`) and named in `groupingFactors`:

```{r}
#| label: boxplot
vals <- matrix(c(5.1, 4.9, 5.5, 6.2, 5.8, 7.1, 6.9, 7.4, 8.2, 7.9, 8.5, 9.1),
               nrow = 1,
               dimnames = list("Measurement", paste0("s", 1:12)))

grp <- data.frame(Group = rep(c("A", "B", "C"), each = 4),
                  row.names = paste0("s", 1:12))

canvasXpress(data = vals,
             smpAnnot        = grp,
             graphType       = "Boxplot",
             title           = "Distribution by Group",
             groupingFactors = list("Group"),
             colorBy         = "Group",
             showLegend      = TRUE)
```

## How it works

- `canvasXpress(...)` returns an htmlwidget, so it renders in the RStudio Viewer, in
  Quarto/R Markdown HTML output, and in Shiny.
- Data is a matrix or data frame; row/column annotations become CanvasXpress `varAnnot` /
  `smpAnnot` for grouping and coloring.
- Because it is a self-contained widget, the library travels with the page — nothing is
  loaded from a CDN.

To render this page:

``` bash
quarto render demo/demo-r.qmd --to html
```

Or open it in **RStudio** and click **Render** (⇧⌘K). See the
[extension README](https://github.com/neuhausi/canvasxpress-quarto) for the JSON-spec
shortcode alternative in [`demo.qmd`](demo.qmd).
